---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%",
  dpi = 150
)

library(ggplot2)

theme_set(theme_bw(base_size = 11))

theme_update(
  legend.position = "none",
  axis.text = element_text(color = "black", size = 11),
  axis.title.x.bottom = element_text(margin = margin(t = 11)),
  axis.title.y.left = element_text(margin = margin(r = 11))
)
```

# lmls

<!-- badges: start -->
[![R CMD check](https://github.com/hriebl/lmls/actions/workflows/check-standard.yaml/badge.svg)](https://github.com/hriebl/lmls/actions)
[![Codecov test coverage](https://codecov.io/gh/hriebl/lmls/branch/main/graph/badge.svg)](https://app.codecov.io/gh/hriebl/lmls?branch=main)
<!-- badges: end -->

*lmls* is an R package for working with linear models for location and scale. These models are multi-predictor regression models with explanatory variables for the mean (= the location) and the standard deviation (= the scale) of a normally distributed response variable. They are special cases of generalized additive models for location, scale and shape (GAMLSS), as described by [Rigby and Stasinopoulos (2005)](https://doi.org/10.1111/j.1467-9876.2005.00510.x).

The *lmls* package provides functions for maximum likelihood and Markov chain Monte Carlo (MCMC) estimation, a parametric bootstrap algorithm, and some diagnostic plots. It was written for the "Advanced Statistical Programming" course at Göttingen University. Feel free to use the package and vignette as an introduction to location-scale regression, or as a basis for the implementation of additional inference algorithms and model extensions.

## Installation

You can install the released version of *lmls* from [CRAN](https://cran.r-project.org/package=lmls) with:

``` r
install.packages("lmls")
```

Or install the development version from GitHub:

``` r
# install.packages("remotes")
remotes::install_github("hriebl/lmls")
```

## Example

The *lmls* package comes with the *abdom* dataset (which it borrows from the *gamlss.data* package). The dataset consists of only two variables: the size of 610 fetuses (as measurements of their abdominal circumference taken from ultrasound scans) and their gestational age ranging from 12 to 42 weeks.

To estimate an LMLS with a quadratic effect of the babies' age on their average size and a linear effect on the log-standard deviation, run this code:

```{r abdom-model}
library(ggplot2)
library(lmls)

m <- lmls(y ~ poly(x, 2), ~ x, data = abdom)
```

You can now use *ggplot2* to visualize the model:

```{r abdom-plot}
df <- abdom
df$mu <- predict(m, type = "response", predictor = "location")
df$sigma <- predict(m, type = "response", predictor = "scale")
df$upper <- df$mu + 1.96 * df$sigma
df$lower <- df$mu - 1.96 * df$sigma

ggplot(df, aes(x, y)) +
  geom_point(color = "darkgray", size = 1) +
  geom_line(aes(y = mu), linewidth = 0.7) +
  geom_line(aes(y = upper), linewidth = 0.3) +
  geom_line(aes(y = lower), linewidth = 0.3) +
  xlab("Age [weeks]") +
  ylab("Size [mm]")
```

## Vignette

For more details, see the [package vignette](https://hriebl.github.io/lmls/articles/lmls.pdf).
